HO unification from object language to meta language

### 2024-04-03 paper/main.tex
notes

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the original unification problem is not in \llambda
  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  \todo[inline]{Is it clearer?}
  and the unification fails.
\end{corollary}

@@

\begin{proof}[Proof sketch]
  By contradiction, if $m = 0$, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.
\end{proof}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  \todo[inline]{Is it clearer?}
\end{definition}

@@

\begin{proof}[Proof sketch]
  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  creates each time a new \linketa and replace the old \linkbeta with a new one.
  This progression can however triggered at most $m$ times, since each new
  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  the \linkbeta is removed from \linkStore which is now filled by $m$ new
  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  terminates if \linkStore is made by only \linketa and therefore
  \elpiIn{progress} still terminates.
  \todo[inline]{is it ok?}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the variables occurring in its \rhs is in \llambda.
  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
  \todo[inline]{is it clearer?}
\end{corollary}

@@

\paragraph{Example of \progBetaLL} 
A simple example of \linkbeta progression due to \progBetaLL is given below:

@@

The \linkbeta in \linkStore has a variable $D$ applied to the variable $E_x$.
The first unification problem, $\hoUnifPb_1$ produces the substitution $\sigma =
\{A \mapsto \lambda x.x\}$. This instantiation triggers $\linkStore_1$ which
before its removal, assignes $E$ to $\lambda x.x$. Now, the \linkbeta is
\linkbetaM{x}{C_x}{(D \appsep x)}. This link is replaced with the couple of
links: \linkbetaM{x}{C_x}{F_x}, \linketaM{}{E}{\lambda x.D_x}. $\hoUnifPb_2$
assignes $B$ which activates $\linkStore_2$, and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto a\}$.

@@

The first unification, $\hoUnifPb_1$, gives the substitution $\sigma = \{A
\mapsto \lambda x.B\}$. The \linkbeta, $\linkStore_1$, becomes
\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
$B$. $B$ is in \llambda and is unified with $C$. The resolution of $\hoUnifPb_2$
gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C
\mapsto a\}$ which is decompiled into $\rho = \{X \mapsto \lambda x.a, Y \mapsto
a\}$.

>>>>

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the original unification problem is not in \llambda
  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  %\todo[inline]{Is it clearer?}
  and the unification fails.
  \todo[inline]{it is clear. Quello che manca è percè serve farlo, e che cosa si guadagna. Qui dimostri che fallire "ha senso" se siamo solo interessati a llambda, ma allo stesso tempo sti beta link sono lì per gestire roba fuori.}
\end{corollary}

@@

\begin{proof}[Proof sketch]
  By contradiction,\todo{we assume to have a link} if $m = 0$, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.\todo{since we started by assuming we had a link}
\end{proof}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  %\todo[inline]{Is it clearer?}
  \todo[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
\end{definition}

@@

\begin{proof}[Proof sketch]
  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  creates each time a new \linketa and replace the old \linkbeta with a new one.
  This progression can however triggered at most $m$ times, since each new
  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  the \linkbeta is removed from \linkStore which is now filled by $m$ new
  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  terminates if \linkStore is made by only \linketa and therefore
  \elpiIn{progress} still terminates.
  %\todo[inline]{is it ok?}
  \todo[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
  \todo[inline]{is it clearer?}
\end{corollary}

@@

\paragraph{Example of \progBetaLL}
Consider the \linkbeta below:

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers progress \todo{which rule?} in first \linketa that is 
removed and $E$ is assigned to $\lambda x.x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the following two links:
\linkbetaM{x}{C_x}{F_x} and
\todo{ma allora il link ora sparisce perhè è in llambda?}
\linketaM{}{E}{\lambda x.D_x}. The second unification
problem assignes $a$ to $B$, that in turn activates
the second \linketa ($a$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto a\}$.\todo{assegnament per D incompleto}
\todo{alla fine non si usa anche la regola beta-rigid-head?}

@@

The first unification problems is solved by the substitution $\sigma = \{A
\mapsto \lambda x.B\}$. The \linkbeta becomes
\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
$B$. $B$ is in \llambda and is unified with $C$. The resolution of the second
unification problem
gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C
\mapsto a\}$ which is decompiled into $\rho = \{X \mapsto \lambda x.a, Y \mapsto
a\}$.
